Deterministic and statistical truncated KdV models for anomalous waves induced by abrupt depth change

### 2018-09-28 LatexNotes/Notes.tex
NM

@@

\np
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% TKDV
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The truncated KdV equations}

\subsection{The KdV equation}

\subsubsection{The constant-depth KdV equation in dimensional form}

The important thing to remember is that Johnson uses these characteristic scales for non-dimensionalizing (characteristic scales denoted by a *)
\begin{align}
\eta^* = \eps h_0 , \quad
x^* = \eps^{-1/2} h_0 , \quad
t^* = \eps^{-1/2} h_0/c
\end{align}
where $h_0$ is a characteristic depth (say the upstream depth), $c = \sqrt{g h_0}$ the corresponding wave speed, and $\eps = a / h_0$ is the ratio of wave amplitude to depth. Roughly speaking, we want to alter Johnson's non-dimensionalization so that $x$ is scaled on some characteristic wavelength (and $t$ scaled accordingly). Importantly, we need to account for variable depth.

\subsubsection{The variable-depth KdV equation}

We start with the variable-depth KdV equation given by Johnson in Eq.(3.149) on page 272 in dimensionless form 
\begin{equation}
\label{varKdVJohnson}
2 H_X + \frac{3}{D^{7/4}} H H_\xi + \frac{1}{3} D^{1/2} H_{\xi \xi \xi} = 0
\end{equation}
All variables above are dimensionless, and by working backwards we can figure out how they relate to the original, dimensional variables. Below, the dimensional quantities are indicated with an overbar wherever there is the possibility of ambiguity
\begin{align}
& H = D^{1/4} \eta = \left( \frac{d}{h_0} \right)^{1/4} \frac{\bar{\eta}}{\eps h_0} \\
& X = \eps x = \frac{\eps^{3/2}}{h_0} \bar{x} \\
& \xi = \frac{\eps^{1/2}}{\sqrt{d h_0}} (\bar{x} - c \bar{t})
\end{align}
Here $d$ is the variable depth (which is dimensional) $c = \sqrt{g d}$ is the local wave speed, and $D = d/h_0$ is the depth ratio. The last relationship above for $\xi$ is valid for the case of two depths (I am not sure if it is valid in the more general case of variable depth).

There are a few important things to remember when we reformulate this equation:
\begin{enumerate}
\item We must retain Johnson's choice to keep the combination $D^{1/4} \eta$ together because that quantity matches across the abrupt depth change (the raw surface displacement does not match).
\item We want to normalize the standard deviation (or energy) of $\eta$ and therefore want to keep Johnson's choice to scale surface displacement on $a = \eps h_0$.
\item We have to scale the spatial variable (called $x$ or $\xi$) on a characteristic wavelength, which will be a different value on either side of the abrupt depth change. We have to do this so that the assumption of spatial periodicity will be reasonably satisfied on both sides.
\end{enumerate}
To achieve 3, we can use the characteristic wavelength from shallow water theory as a guide,
\begin{equation}
\lambda_c = {c}/{f_c} = \sqrt{g d}/f_c
\end{equation}
where $f_c$ is a characteristic forcing frequency and $c = \sqrt{g d}$ is the local wavespeed (in shallow-water theory). It is tempting to simply set $f_c$ to the peak forcing frequency $f_p$, which would result in the characteristic wavelength $\lambda_c$ being the peak wavelength $\lambda_p$. Perhaps a better choice, though, is to let $\lambda_c$ be an integer multiple of $\lambda_p$ to allow scales larger than the peak wavelength (which are certainly present in the experiments). We therefore choose
\begin{equation}
\lambda_c = M \lambda_p = M \sqrt{gd} / f_p
\end{equation}
where $M$ is an integer (on the order of 10 or so). Notice that Johnson's formulation already incorporates the correct scaling of $\sqrt{d}$ to define $\xi$, which will make things convenient. Thus, the main change we need to make is to introduce the dimensionless variable
\begin{equation}
\zeta = \frac{2\pi}{\lambda_c} (\bar{x} - c \bar{t}) 
= \frac{2 \pi \delta}{M \eps^{1/2}} \, \xi
\end{equation}
where
\begin{equation}
\delta = h_0 / \lambda_0 = f_p \sqrt{h_0/g}
\end{equation}
To clean things up a little, lets also introduce a slightly different long-space coordinate that uses the upstream wavelength $\lambda_0$ as a characteristic scale instead of $h_0$,
\begin{equation}
s = \frac{2 \pi \beta}{M \lambda_0} \bar{x} = 2 \pi \beta M^{-1} \delta \eps^{-3/2} X
\end{equation}
where $\beta$ is a small parameter that can be taken to be, for example, $\eps$ or $\delta^2$ or something else.

\subsubsection{Final formulation}
Equation \eqref{varKdVugly} is essentially in the form we want to use, except that the notation is bad. In particular, we will want to reserve the symbol $H$ for the Hamiltonian. So making superficial changes to the notation gives
\begin{align}
\label{varKdV1}
2 \beta u_s + 3 \eps D^{-7/4} \, u u_x
+ \frac{4 \pi^2 \delta^2}{3 M^2} D^{1/2} \, u_{xxx} = 0
\end{align}
Importantly, $u(x,s)$ is $2\pi$-periodic in $x$ and has normalized energy $\mean{u^2} = 1$.
The definitions of each of the dimensionless variables in terms of the original, physical ones (denoted with a bar) are
\begin{align}
& u = D^{1/4} \, \bar{\eta}/a \\
& x = 2\pi \lambda_c^{-1} (\bar{x} - c \bar{t})  \\
& s = 2 \pi \beta M^{-1} \lambda_0^{-1} \bar{x}
\end{align}
A list of the fixed parameters is
\begin{align}
& \lambda_0 = \sqrt{g h_0}/f_p \quad \text{(upstream wavelength)} \\
& \eps = a / h_0 \quad \text{(dimensionless ratio of wave amplitude to depth)} \\
& \delta = h_0 / \lambda_0 \quad \text{(dimensionless ratio of depth to upstream wavelength)} \\
& \beta	\quad \text{(an unspecified small parameter, usually $\eps$ or $\delta^2$)}
\end{align}
Here is a list of the parameters that vary in space due to the fact that depth $d$ varies in space
\begin{align}
& c = \sqrt{gd} \quad \text{(the wavespeed as it varies in space)} \\
&\lambda_c = M \lambda_p = M \sqrt{gd}/f_p \quad \text{(characteristic wavelength as it varies in space)} \\
&D = d/h_0 \quad \text{(dimensionless depth ratio as it varies in space)} \\	
\end{align}
Note, in the experiments, $\delta = 0.33$ and $\eps$ lies in the range $2.4 \times 10^{-3}$--$2.4 \times 10^{-2}$.

>>>>

 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% KdV
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The KdV equations}

\subsection{The constant-depth KdV equation in dimensional form}

The important thing to remember is that Johnson uses these characteristic scales for non-dimensionalizing (characteristic scales denoted by a *)
\begin{align}
\eta^* = \eps h_0 , \quad
x^* = \eps^{-1/2} h_0 , \quad
t^* = \eps^{-1/2} h_0/c
\end{align}
where $h_0$ is a characteristic depth (say the upstream depth), $c = \sqrt{g h_0}$ the corresponding wave speed, and $\eps = a / h_0$ is the ratio of wave amplitude to depth. We want to alter Johnson's non-dimensionalization so that $x$ is scaled on some characteristic wavelength (and $t$ scaled accordingly). Importantly, we need to account for variable depth.

\subsection{The variable-depth KdV equation}

We start with the variable-depth KdV equation given by Johnson in Eq.(3.149) on page 272 in dimensionless form 
\begin{equation}
\label{varKdVJohnson}
2 H_X + \frac{3}{D^{7/4}} H H_\xi + \frac{1}{3} D^{1/2} H_{\xi \xi \xi} = 0
\end{equation}
All variables above are dimensionless, and by working backwards we can figure out how they relate to the original, dimensional variables. Below, the dimensional quantities are indicated with an overbar wherever there is the possibility of ambiguity
\begin{align}
& H = D^{1/4} \eta = \left( \frac{d}{h_0} \right)^{1/4} \frac{\bar{\eta}}{\eps h_0} \\
& X = \eps x = \frac{\eps^{3/2}}{h_0} \bar{x} \\
& \xi = \frac{\eps^{1/2}}{\sqrt{d h_0}} (\bar{x} - c \bar{t})
\end{align}
Here $d$ is the variable depth (which is dimensional) $c = \sqrt{g d}$ is the local wave speed, and $D = d/h_0$ is the depth ratio. The last relationship above for $\xi$ is valid for the case of two depths (though perhaps not in the general case).

There are a few important things to remember when we reformulate this equation:
\begin{enumerate}
\item We must retain Johnson's choice to keep the combination $D^{1/4} \eta$ together because that quantity matches across the abrupt depth change (the raw surface displacement does not match).
\item We want to normalize the standard deviation (or energy) of $\eta$ and therefore want to keep Johnson's choice to scale surface displacement on $a = \eps h_0$.
\item We have to scale the spatial variable (called $x$ or $\xi$) on a characteristic wavelength, which will be a different value on either side of the abrupt depth change. We have to do this so that the assumption of spatial periodicity will be reasonably satisfied on both sides.
\end{enumerate}
To achieve 3, we can use the characteristic wavelength from shallow water theory,
\begin{equation}
\lambda_c = {c}/{f_c} = \sqrt{g d}/f_c
\end{equation}
where $f_c$ is a characteristic forcing frequency and $c = \sqrt{g d}$ is the local wavespeed. It is tempting to simply set $f_c$ to the peak forcing frequency $f_p$, which would result in the characteristic wavelength $\lambda_c$ being the peak wavelength $\lambda_p$. Perhaps a better choice, though, is to let $\lambda_c$ be an integer multiple of $\lambda_p$ to allow scales larger than the peak wavelength (which are certainly present in the experiments). We therefore choose
\begin{equation}
\lambda_c = M \lambda_p = M \sqrt{gd} / f_p
\end{equation}
where $M$ is an integer (on the order of 10 or so). Notice that Johnson's formulation already incorporates the correct scaling of $\sqrt{d}$ to define $\xi$, which will make things convenient. Thus, the main change we need to make is to introduce the dimensionless variable
\begin{equation}
\zeta = \frac{2\pi}{\lambda_c} (\bar{x} - c \bar{t}) 
= \frac{2 \pi \delta}{M \eps^{1/2}} \, \xi
\end{equation}
where
\begin{equation}
\delta = h_0 / \lambda_0 = f_p \sqrt{h_0/g}
\end{equation}
To clean things up a little, lets also introduce a slightly different long-space coordinate that uses the upstream wavelength $\lambda_0$ as a characteristic scale instead of $h_0$,
\begin{equation}
s = \frac{2 \pi \beta}{M \lambda_0} \bar{x} = 2 \pi \beta M^{-1} \delta \eps^{-3/2} X
\end{equation}
where $\beta$ is a small parameter that can be taken to be, for example, $\eps$ or $\delta^2$ or something else.

\subsection{Final formulation}
Equation \eqref{varKdVugly} is essentially in the form we want to use, except that the notation is bad. In particular, we will want to reserve the symbol $H$ for the Hamiltonian. So making superficial changes to the notation gives
\begin{align}
\label{varKdV1}
2 \beta u_s + 3 \eps D^{-7/4} \, u u_x
+ \frac{4 \pi^2 \delta^2}{3 M^2} D^{1/2} \, u_{xxx} = 0
\end{align}
Importantly, $u(x,s)$ is $2\pi$-periodic in $x$ and has normalized energy $\mean{u^2} = 1$.
The definitions of each of the dimensionless variables in terms of the original, physical ones (denoted with a bar) are
\begin{align}
& u = D^{1/4} \, \bar{\eta}/a \\
& x = 2\pi \lambda_c^{-1} (\bar{x} - c \bar{t})  \\
& s = 2 \pi \beta M^{-1} \lambda_0^{-1} \bar{x}
\end{align}
A list of the fixed parameters is
\begin{align}
& \lambda_0 = \sqrt{g h_0}/f_p \quad \text{(upstream wavelength)} \\
& \eps = a / h_0 \quad \text{(dimensionless ratio of wave amplitude to depth)} \\
& \delta = h_0 / \lambda_0 \quad \text{(dimensionless ratio of depth to upstream wavelength)} \\
& \beta	\quad \text{(an unspecified small parameter, usually $\eps$ or $\delta^2$)}
\end{align}
Here is a list of the parameters that vary in space due to the fact that depth $d$ varies in space
\begin{align}
& c = \sqrt{gd} \quad \text{(the wavespeed as it varies in space)} \\
&\lambda_c = M \lambda_p = M \sqrt{gd}/f_p \quad \text{(characteristic wavelength as it varies in space)} \\
&D = d/h_0 \quad \text{(dimensionless depth ratio as it varies in space)} \\	
\end{align}
Note, in the experiments, $\delta = 0.33$ and $\eps$ lies in the range $2.4 \times 10^{-3}$--$2.4 \times 10^{-2}$.

Note: $\eps \delta^{-2} = a f_p^2 / g$ which is an interesting parameter. It suggests an equivalence between driving amplitude and driving frequency.

@@
